---
title: "15 Minute Cities in the Urban-Rural Nexus"
output: html_notebook
---

This tutorial shows one way to compute and map the 15 minute city with R, focusing on Urban Rural differences.


**Access to Services based on Euclidean Distance**

As a first step, services that can be accessed in 10 minute walk need to be identified. To do so, we will use a metric based on Euclidean distance, selecting POIs which fall within a 800 meters radius from each postcode in the Lothian region.

```{r}
library(tidyverse) # this library will be used for all data wrangling instructions
library(sf) # this is a well-established library to deal with spatial data
library(nngeo) #this library will be used to compute distances 

#DO NOT RUN 

#load the data and pre-process

#origins<-st_read("data/postcodes.gpkg") %>% #UK data have a specific Coordinate Reference System (British National Grid)
#  st_transform(4326) #globally WGS84 is generally used. Here, we therefore transform the CRS to WGS84 

#origins <- origins %>% #create a unique ID for the origins
#  mutate(
#    id=as.numeric(rownames(origins))
#  )

#destinations<-st_read("data/POIs.gpkg") 

#destinations<- destinations %>% #create a unique ID for the destinations
#  mutate(
#    id=as.numeric(rownames(destinations))
#  )

#Compute distances from each origin and select POIs within 800 meters 

#nn_out <- st_nn(origins, destinations, k=50, maxdist = 800) 

#The steps above take too much time and are heavy, for the purpose of this tutorial you will directly load the results. You will have access to the fully reproducible repository that you can re-use with your machine
nn_out <- readRDS("data/nn_out.RDS")

#post-processing
#transform results from list to dataframe
results <- data.frame(
              id = rep(
              seq_along(nn_out),
              lengths(nn_out)
              ),
              dest = unlist(nn_out)
            ) %>%
          inner_join(st_drop_geometry(origins), by = c("id" = "id"))%>%
          inner_join(st_drop_geometry(destinations), by = c("dest" = "id"))

head(results)

#compute the average number of poi accessible in 10-min walk in each data zone and the average number of services type

results_processed <- results %>%
  group_by(DataZone) %>%
  summarise(
    poi = n(),
    service=length(unique(fclass)),
    pc = length(unique(id)),
    poiAvg = poi/pc
    )

head(results_processed)


```

**Mapping the Results**

We can now visualise the results on a map to reflect on the spatial distribution of services.

```{r}
library(ggplot2)

#load spatial boundaries and join them with the results obtained in the previous section
#let's start with the number of different service by data zone
dataZones<-st_read("data/dataZones_UR.gpkg") %>%
  inner_join(results_processed, by=c("DataZone"="DataZone"))

colnames(dataZones)

#mapping the average poi accessible in 10 minute walk per resident (pp)

ggplot()+
  geom_sf(aes(fill = poiAvg/(ResPop2011/pc)),
          colour = "NA",
          data = dataZones) +
  #change color palette
  scale_fill_viridis_c() +
  #change the background
  theme_void() + 
  #add a scale bar
  ggspatial::annotation_scale(  
    location = "bl",
    bar_cols = c("grey60", "white")
    ) +
  #change legend title
  labs(fill = "POI Accessible in 10 min walk\n(Average pp)")+
  #customize theme
  theme(
    text = element_text(family = "Futura-Medium",color = "#22211d"),
    legend.title = element_text(family = "Futura-Bold", size = 10),
    legend.text = element_text(family = "Futura-Medium", size = 10),
    plot.background = element_rect(fill = "#f5f5f2", color = NA),
    panel.background = element_rect(fill = "#f5f5f2", color = NA)
    )  


ggplot()+
  geom_sf(aes(fill = service),
          colour = "NA",
          data = dataZones) +
  #change color palette
  scale_fill_viridis_c() +
  #change the background
  theme_void() + 
  #add a scale bar
  ggspatial::annotation_scale(  
    location = "bl",
    bar_cols = c("grey60", "white")
    ) +
  #change legend title
  labs(fill = "Service's Types\naccessible in 10 min walk")+
  #customize theme
  theme(
    text = element_text(family = "Futura-Medium",color = "#22211d"),
    legend.title = element_text(family = "Futura-Bold", size = 10),
    legend.text = element_text(family = "Futura-Medium", size = 10),
    plot.background = element_rect(fill = "#f5f5f2", color = NA),
    panel.background = element_rect(fill = "#f5f5f2", color = NA)
    )  

#create a score which accounts for both POIs per person and service's types

dataZones<-dataZones %>%
  mutate(
    score = scale(poiAvg/(ResPop2011/pc)) + scale(service)
    )

#map the score

ggplot()+
  geom_sf(aes(fill = score),
          colour = "NA",
          data = dataZones) +
  #change color palette
  scale_fill_viridis_c() +
  #change the background
  theme_void() + 
  #add a scale bar
  ggspatial::annotation_scale(  
    location = "bl",
    bar_cols = c("grey60", "white")
    ) +
  #change legend title
  labs(fill = "Access Score")+
  #customize theme
  theme(
    text = element_text(family = "Futura-Medium",color = "#22211d"),
    legend.title = element_text(family = "Futura-Bold", size = 10),
    legend.text = element_text(family = "Futura-Medium", size = 10),
    plot.background = element_rect(fill = "#f5f5f2", color = NA),
    panel.background = element_rect(fill = "#f5f5f2", color = NA)
    )  


```


**The Rural-Urban nexus**
Let's now look at the gap between access to services between urban and rural areas.

The Scottish Urban-Rural classification distinguishes between:

1) Large Urban Areas 	Settlements of 125,000 people and over.

2) Other Urban Areas 	Settlements of 10,000 to 124,999 people.

3) Accessible Small Towns 	Settlements of 3,000 to 9,999 people, and within a 30 minute drive time of a Settlement of 10,000 or more.

4) Remote Small Towns 	Settlements of 3,000 to 9,999 people, and with a drive time of over 30 minutes to a Settlement of 10,000 or more.

5) Accessible Rural Areas 	Areas with a population of less than 3,000 people, and within a 30 minute drive time of a Settlement of 10,000 or more.

6) Remote Rural Areas 	Areas with a population of less than 3,000 people, and with a drive time of over 30 minutes to a Settlement of 10,000 or more.

```{r}

head(dataZones)
#URValue corresponds to the classes above. Let's create more legible classes

dataZones <- dataZones %>%
  mutate(
    URclasses = case_when(
      URValue == 1 ~ "Large Urban Areas",
      URValue == 2 ~ "Other Urban Areas",
      URValue == 3 ~ "Small Towns",
      URValue == 5 ~ "Accessible Rural Areas"
    ),
    URclasses = factor(URclasses, levels = c("Large Urban Areas", "Other Urban Areas", "Small Towns", "Accessible Rural Areas"))
  )

ggplot()+
  geom_sf(aes(fill = URclasses),
          colour = "NA",
          data = dataZones) +
  #change color palette
  scale_fill_viridis_d() +
  #change the background
  theme_void() + 
  #add a scale bar
  ggspatial::annotation_scale(  
    location = "bl",
    bar_cols = c("grey60", "white")
    ) +
  #change legend title
  labs(fill = "Urban-Rural Classification")+
  #customize theme
  theme(
    text = element_text(family = "Futura-Medium",color = "#22211d"),
    legend.text = element_text(family = "Futura-Medium", size = 10),
    plot.background = element_rect(fill = "#f5f5f2", color = NA),
    panel.background = element_rect(fill = "#f5f5f2", color = NA)
    )  


dataZones %>%
  st_drop_geometry() %>%
  group_by(URclasses) %>%
  summarise(
    c_poiAvg = mean(poiAvg, rm.na = TRUE),
    c_service = mean(service, rm.na = TRUE),
    c_score = mean(score, rm.na = TRUE)
  ) %>%
  pivot_longer(cols = starts_with("c_"), names_to = "variable", values_to = "value")%>%
  mutate(
    variable = factor(variable, levels = c("c_poiAvg","c_service","c_score"), labels = c("Average POIs", "Average Services", "Access Score"))
  ) %>%
  ggplot(aes(x = URclasses, y = value, fill = URclasses))+
  geom_bar(stat = "identity", width = 0.5) +
  scale_fill_viridis_d() +
  facet_wrap(~variable, scales = "free", ncol = 1) +
  labs(fill = "Urban-Rural Classification") +
  theme_minimal()+
  theme(axis.title = element_blank(), axis.text.x = element_blank(), text = element_text(size = 12))
  

```

    